PLAE enables quick and easy searching across one million ocular transcriptomes

### 2022-06-17 heatmaps.Rmd

# Rpe volcano
```{r, fig.width=3, fig.height=2}
rpe_diff <- scEiaD_2020_v01 %>% 
  tbl("diff_testing") %>% 
  filter(Base == 'RPE') %>%  
  filter(Group == 'CellType (Predict)', 
         Organism == org, Against == 'All') %>% 
  select(Gene, log2FoldChange, pvalue, Base) %>% 
  collect()

@@

rpe_diff %>% 
  ggplot(aes(x=log2FoldChange,y=-log2(pvalue))) + 
  geom_point() + cowplot::theme_cowplot() + ggrepel::geom_label_repel(data = rpe_diff %>% 
                                                                        filter(Gene %in% genes) %>% 
                                                                        mutate(Gene = gsub('\\(|\\)|ENSG\\d+','',Gene) %>% 
                                                                                 gsub(' ','',.)), 
                                                                      aes(label = Gene),
                                                                      max.overlaps = 100)
```

>>>>

# Rpe volcano
```{r, fig.width=3, fig.height=2}
rpe_diff <- scEiaD_2020_v01 %>% 
  tbl("diff_testing") %>% 
  filter(Base == 'RPC') %>%  
  filter(Group == 'CellType (Predict)', 
         Organism == org, Against == 'All') %>% 
  select(Gene, log2FoldChange, pvalue, Base, baseMean) %>% 
  collect()

@@

rpe_diff %>% 
  ggplot(aes(x=log2FoldChange,y=-log2(pvalue))) + 
  geom_point(aes(size=log2(baseMean))) + cowplot::theme_cowplot() + ggrepel::geom_label_repel(data = rpe_diff %>% 
                                                                        filter(Gene %in% genes) %>% 
                                                                        mutate(Gene = gsub('\\(|\\)|ENSG\\d+','',Gene) %>% 
                                                                                 gsub(' ','',.)), 
                                                                      aes(label = Gene),
                                                                      max.overlaps = 100)
```

### 2022-06-17 manuscript_00.Rmd

## Differential testing identifies thousands genes enriched across various ocular cell types

We used a "pseudo bulk" based approach wherein the counts for a category (CellType, Cluster, or CellType (Predict)) were summed for each category - organism - study grouping. The creates a counts matrix which has statistical properties matching that of "bulk" RNA seq experiment matrices. This approach has been found to be more specific in identifying differentially expressed genes in single cell RNA-seq experiments [@squairConfrontingFalseDiscoveries2021]. Once the aggregated matrix was constructed, we use the DESeq2 differential gene expression analysis software to identify differentially expressed genes. We tested each species (Human, Macaque, Mouse) separately. We had an insufficient number of chicken studies to conduct differential testing. The model used for DESeq2 was "~study_accession + category" in order to regress out study specific differential expression. After running these tests we extracted differentially expressed genes with two types of contrasts: either against all remaining categories (e.g. cones versus all remaining cell types) or in a pair-wise manner (cone against rod). We find
```{r}
# fix  occasional NA
# data made from src/ensdb.R
diff_tab_big <- read_tsv('data/diff_tab_big.tsv.gz')
gene_tab <- bind_rows(read_tsv('data/hs.ensdb.v86.symbol.geneid.tsv.gz'), read_tsv('data/mm.ensdb.v79.symbol.geneid.tsv.gz'))
diff_tab <- diff_tab_big %>% 
  filter(Against == 'All', Base %in% (meta_filter %>% filter(Organ == 'Eye') %>% pull(CellType) %>% unique()))

>>>>

## Leveraging the high diversity of studies to identify thousands of consistently differentially expressed genes across ocular cell types

We used a "pseudo-bulk" based approach where the counts for a category (CellType, Cluster, or CellType (Predict)) were summed for each category - organism - study grouping. The created a counts matrix which has statistical properties matching that of "bulk" RNA seq experiment matrices and thus allowing for mature bulk RNA-seq tooling to be used. This approach has been found to be more specific in identifying differentially expressed genes in single cell RNA-seq experiments [@squairConfrontingFalseDiscoveries2021]. Once the aggregated matrix was con structured, we use the DESeq2 differential gene expression analysis software to identify differentially expressed genes. We tested each species (Human, Macaque, Mouse) separately. We had an insufficient number of chick studies to conduct differential testing. The model used for DESeq2 was "~study_accession + category" in order to regress out study specific differential gene expression. After running these tests we extracted differentially expressed genes with two types of contrasts: either against all remaining categories (e.g. cones versus all remaining cell types) or in a pair-wise manner (cone against rod). We find, in total, thousands of differentially expressed genes across the ocular cell types.

```{r}
# fix  occasional NA
# data made from src/ensdb.R
diff_tab_big <- read_tsv('data/diff_tab_big.tsv.gz')
gene_tab <- bind_rows(read_tsv('data/hs.ensdb.v86.symbol.geneid.tsv.gz'), read_tsv('data/mm.ensdb.v79.symbol.geneid.tsv.gz'))
diff_tab <- diff_tab_big %>% 
  filter(Against == 'All', Base %in% (meta_filter %>% filter(Organ == 'Eye') %>% pull(CellType) %>% unique()))
